A SAT-Based Counterexample Guided Method for Unbounded Synthesis

### 2016-01-25 paper.tex
more editing + optimisations section

The entire set of runs of a bounded game of rank $n$ is
symbolically represented by a tree of depth $2n$ and width 1 populated by
unfixed edges.  Reducing the set of runs in a game forms an abstract game,
which can be represented symbolically by an \emph{abstract game tree}.

@@

An abstract game tree can be checked for the existence of a winning run by a
SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
copies of the transition relation for each game step in tree. The formula must
also check whether the error state has been reached in no branch/in all
branches for the controller and environment respectively. The functions that
produce these formulas are shown in Algorithm \ref{alg:treeFormula}.

@@

When we produce a formula using $\textsc{treeFormula}$ the SAT solver is
playing on behalf on the controller. Any unfixed labels in the tree,
controllable or uncontrollable, will be existentially quantified.  This means
that if there exists a way for the game to be solved when both players
cooperate the SAT solver will find it. If no winning run exists in an abstract
game even when the players are cooperating then there is definitely no winning
run when the opponent is playing adversarily.

@@

By checking for the existence of a winning run for the environment in an
abstract game tree constructed from a partial controller strategy we are
checking if the partial strategy is winning. If no spoiling run can be found
then the partial strategy must always win. If a spoiling run is found then we
then we have a counterexample that proves that the controller strategy is not
winning. This forms the basis of a counterexample guided abstraction refinement
framework that operates on candidate strategies.

This technique was proposed by Narodytska et al \cite{narodytska2014}. In
Section \ref{sect:unbounded} we propose an extension to synthesis of unbounded
games.

The bounded synthesis algorithm begins with an empty game as seen in Figure
\ref{fig:emptytree}.  Initially we are playing on behalf of the environment
because it chooses the first move in each step.  The empty game is passed to
the SAT solver, which searches for a candidate environment strategy.  If a
candidate is found (Figure \ref{fig:candidate}) then it is checked for a
spoiling strategy by solving for the controller.  If no spoiling strategy
exists, that means our candidate is a winning strategy and the algorithm
terminates.  Otherwise we find a counterexample (Figure \ref{fig:cex}), which
is used to refine the empty game tree to include the first move from the
controller's spoiling strategy (Figure \ref{fig:refined}.  The algorithm
continues by finding a new candidate for the environment (Figure
\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
once again (Figure \ref{fig:refined2}). The full algorithm is listed in
Algorithm \ref{alg:bounded}.

@@

Bounded synthesis can be used to prove the existence of a winning strategy for
the environment by providing a witness. For the controller, the strongest claim
that can be made is that the strategy is winning as long as the game does not
extend longer than the bound.

In model checking the maximum bound is decided based on the states of the game
\cite{biere1999}. The na\"ive approach is to use size of the state space as the
bound ($2^{|X|}$). With a bound of this size all states may be explored. One
optimisation is to use the diameter of the game, which is the smallest number
$d$ such that for any state $x$ there is a path of length $\leq d$ to all other
reachable states. However, for large games these bounds are intractable.

@@

Given two formulas $A$ and $B$ such that $A \land B$ is unsatisfiable, it is
possible to construct and interpolant $\mathcal{I}$ such that $A \to
\mathcal{I}$, $B \land \mathcal{I}$ is unsatisfiable, and $\mathcal{I}$ refers
only to the intersection of variables in $A$ and $B$. An interpolant can be
constructed efficiently from a resolution proof of the unsatisfiability of $A
\land B$ \cite{pudlak1997}.

@@

Consider the snippet of a game tree in Figure \ref{fig:interpolatetree}. The
tree is losing for the controller, the node labelled $x_1$ is at rank 1, and
$x_0^0$ and $x_0^1$ are at rank 0. Since the tree is controller-losing we know
that at least one run represented by the tree contains the error state.  As
a result, $\textsc{treeFormula}(gt)$ is unsatisfiable. If we take the step
$x_1$ to $x_0^0$ and cut it from the rest of the tree then
$\textsc{treeFormula}(step) \land \textsc{treeFormula}(parent)$ must still be
unsatisfiable.

@@

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment respectively}
    \label{alg:treeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf2}
    \end{algorithmic}

@@

In this section we present some optimisations to the unbounded synthesis algorithm.

@@

>>>>

The entire set of runs of a bounded game of rank $n$ is symbolically
represented by a tree of depth $2n$ (a step is two moves - environment then
controller) and width 1 populated by unfixed edges.  Fixing some moves and
thereby reducing the set of runs forms an \emph{abstract game tree} that
symbolically represents a smaller version of the game.

@@

An abstract game tree can be checked for the existence of a winning run by a
SAT solver\cite{narodytska2014}.  The tree must be encoded into CNF by making
copies of the transition relation for each game step in tree. The formula must
also check whether the error state has been reached in no branch
(\emph{respectively}: in all branches) for the controller (\emph{resp}:
environment.) The functions that produce these formulas are shown in Algorithm
\ref{alg:treeFormula}.

@@

When we produce a formula using $\textsc{treeFormula}$ the SAT solver is
playing on behalf on the controller. Any unfixed labels in the tree,
controllable or uncontrollable, will be existentially quantified.  This means
that if there exists a way for the game to be solved when both players
cooperate on unfixed labels the SAT solver will find it. If no winning run
exists in an abstract game even when the players are cooperating then there is
no winning run when the opponent is playing adversarily.

@@

By checking for the existence of a winning run for the environment in an
abstract game tree constructed from a partial controller strategy we are
checking if the partial strategy is winning. If no environment-winning run is
exists within the partial strategy then it is a winning partial strategy,
otherwise the spoiling run is a counterexample that proves that the partial
strategy is not winning. This forms the basis of a counterexample guided
abstraction refinement framework that operates on candidate strategies.

This technique was proposed by Narodytska et al \cite{narodytska2014}. In
Section \ref{sect:unbounded} we propose an extension to synthesis of unbounded
games but the original algorithm is outlined here for completeness.

The bounded synthesis algorithm begins with an empty game as seen in Figure
\ref{fig:emptytree}.  Initially we are playing on behalf of the environment
because it chooses the first move in each step.  The empty game is passed to
the SAT solver, which searches for a candidate environment strategy.  If a
candidate is found (Figure \ref{fig:candidate}) then it is checked for a
spoiling strategy by solving for the controller.  If no spoiling strategy
exists, that means our candidate is a winning strategy and the algorithm
terminates.  Otherwise we find a counterexample (Figure \ref{fig:cex}), which
is used to refine the empty game tree to include the first move from the
controller's spoiling strategy (Figure \ref{fig:refined}).  The algorithm
continues by finding a new candidate for the environment (Figure
\ref{fig:candidate2}), and a new counterexample to refine the game abstraction
once again (Figure \ref{fig:refined2}). The full procedure is listed in
Algorithm \ref{alg:bounded}.

@@

Bounded synthesis can be used to prove the existence of a winning strategy for
the environment on the unbounded game by providing a witness. For the
controller, the strongest claim that can be made is that the strategy is
winning as long as the game does not extend beyond the bound.

It is possible to set a bound such that all runs in the unbounded game will be
considered. The na\"ive approach is to use size of the state space as the bound
($2^{|X|}$) so that all states may be explored by the algorithm. A more nuanced
approach is to use the diameter of the game \cite{biere1999}, which is the
smallest number $d$ such that for any state $x$ there is a path of length $\leq
d$ to all other reachable states. This approach is common in bounded model
checking but for synthesis it quickly becomes intractable to consider such long
runs.

@@

Given two formulas $A$ and $B$ such that $A \land B$ is unsatisfiable, it is
possible to construct a Craig interpolant\cite{craig1957} $\mathcal{I}$ such
that $A \to \mathcal{I}$, $B \land \mathcal{I}$ is unsatisfiable, and
$\mathcal{I}$ refers only to the intersection of variables in $A$ and $B$.  An
interpolant can be constructed efficiently from a resolution proof of the
unsatisfiability of $A \land B$ \cite{pudlak1997}.

@@

Consider the snippet of a game tree in Figure \ref{fig:interpolatetree}. The
tree is losing for the controller, the node labelled $x_1$ is at rank 1, and
$x_0^0$ and $x_0^1$ are at rank 0. Since the tree is controller-losing we know
that at least one run represented by the tree contains the error state.  As
a result, $\textsc{treeFormula}(gt)$ is unsatisfiable. If we take the step
$x_1$ to $x_0^0$ and cut it from the rest of the tree then
$\textsc{treeFormula}(step) \land \textsc{treeFormula}(parent)$ must too be
unsatisfiable.

@@

\begin{algorithm}
    \caption{Amended tree formulas for Controller and Environment respectively}
    \label{alg:unboundedTreeFormula}
    \begin{algorithmic}[1]
        \Function{treeFormula}{gt}
        \If{$\Call{rank}{gt} = 0$}
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}}$ }
        \Else
        \State \Return{ $\lnot \Call{$B^M$}{x^{gt}} \land \bigwedge_{n \in \Call{succ}{gt}}(\Call{$\delta$}{n} \land \Call{label}{n} \land \Call{treeFormula}{n})$ }
        \EndIf
        \EndFunction
        \algstore{tf2}
    \end{algorithmic}

@@

%%%In this section two optimisations to the unbounded synthesis algorithm are
%%%described.

@@

One aspect of the algorithm proposed by Narodytska et al\cite{narodytska2014}
is that it mimics real-world uses of synthesis by allowing cooperation between
system and environment. Unfortunately, allowing the opponents to pick moves for
one another can frequently backfire. It is common when modelling real-world
systems to abstract over out of scope failures with transitions that
immediately determine the outcome of the game. For example, in a network driver
a request to send a packet may fail due to a disconnected wire, which might be
modelled as an environment controlled transition to a system-winning state.

This slows down the synthesis process as all erroneous transition are
accumulated into the abstract game tree. To mitigate this effect we have a
heuristic to guess non-erroneous labels as \emph{default moves} that are used
instead of allowing the opponent to choose. This does not effect the
correctness of the algorithm. If line~(28) previously returned \texttt{sat} and
is \texttt{unsat} with default moves then those moves would eventually have
been added to the abstraction in a refinement step. If the formula is
\texttt{unsat} even with the opponent choosing moves then it must be
\texttt{unsat} with default moves.
